---
title: "DATA624-Project1"
author: "Silma"
date: "2025-03-24"
output: html_document
---

# DATA 624 - Predective Analytics
## Project 1
### Silma Khan
#### Spring 2025

As the industries continue to utilize data for deriving business insights, forecasting has become a crutial addition in aiding business initiatives and planning. This project shows a complete design and workflow for time series forecasting using two different scenarios. We will be covering data import, cleaning, exploration, analysis, model fitting, forecasting, and exporting our results. For this project, it is divided into two parts:

- **Part A - Forecasting ATM Cash Withdrawls**

In this part, we were provided an excel file titled `ATM624Data.xlsx` where we can analyze daily cash withdrawls from 4 different ATM machines. This file contains columns:

1. **DATE**: Where the information if provided in a date and timestamp, which we will convert into a normal date format (maybe create another column for time stamp)
2. **ATM**: where this is a categorical feature than indicates which ATM machine the withdrawl is pulling from. There are `ATM1`, `ATM2`, `ATM3`, and `ATM4`
3. **CASH**: and this contains the information regarding the amount of cash that is withdrawn on that date and from that row specific atm machine. Important to note, the cash is expressed in **hundreds of dollars**, so for example if the column contains 55, then the actual amount withdrawn is $5,500

The goal of Part A is to create a forecast of the daily cash withdrawls for each ATM machine for the entire month of May 2010, as the dates provided stop at April 2010. This will require transforming the data into a time series format where we will have one series per each ATM machine which will then allow us to forecast using a model and then predict the daily withdrawl amounts for the month of May 2010. 

- **Part B - Forecasting Residential Power Usage**

In this part, we were provided with another excel file titled `ResidentialCustomerForecastLoad-624.xlsx` where we will now be using the monthly residential power usage data for years 1998-2013. This file contains the columns:

1. **CaseSequence**: which I believe is used as an identifier which is not necessary to use in the forecast
2. **YYY-MMM**: where this column contails the year and month in the format of 1998-Jan. This will need to go through some transformations and converting it into a standard date format
3. **KWH**: this column contains the power consumption for that month and year in the power unit kilowatt/hours

The goal of Part B is to forecast the power consumption for the months of 2014 since this dataset showcases data from Jan 1998 - Dec 2013. To do this, we would need to convert the date format into a standard date object, create a monthly time series, fitting a forecasting model, and finally generate the forecast for 2014.

First we need to load in our packages we need to perform this project:

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(forecast)
library(openxlsx)
library(ggplot2)
library(zoo)
```
## Part 1 - ATM Cash Forcast

### Data Import

```{r}
getwd()
```

```{r}
atm_data <- read_excel("ATM624Data.xlsx")

head(atm_data)
```



### Data Preprocessing

```{r}
str(atm_data$DATE)
```

```{r}
atm_data$DATE <- as.Date(atm_data$DATE, origin = "1899-12-30")

head(atm_data$DATE)

head(atm_data)
```

After loading the data, we also converted the Excel dates to a proper date object for easier analysis and forecasting

Now, since the ATM machines are all included into one column, I think it would be better to create a wide table format to analyze each ATM component before doing that, we should also check to see if there are any rows with missing ATM values

```{r}
unique(atm_data$ATM)
```
As we can see, NA is part of the count meaning there are some rows that have missing ATM information. To deal with this, we can simply remove rows with missing ATM values

```{r}
atm_data_clean <- atm_data %>%
  filter(!is.na(ATM))
```

After cleaning the data, we can now move onto creating a wide format for the table

```{r}
atm_wide <- atm_data_clean %>%
  select(DATE, ATM, Cash) %>%
  pivot_wider(names_from = ATM, values_from = Cash)

head(atm_wide)
```

Now each row represents a single date and can see columns for the separate ATM machines with their respective cash values withdrawn

Since the date starts from May 2009 and we would like to forecast the cash withdraws from May 2010, we can convert each ATM's cash data into a daily time series

```{r}
atm_wide <- atm_wide %>% arrange(DATE)

create_daily_ts <- function(x, start_date) {
  start_year <- year(start_date)
  start_doy <- yday(start_date)
  ts(x, start = c(start_year, start_doy), frequency = 365)
}

ts_ATM1 <- create_daily_ts(atm_wide$ATM1, min(atm_wide$DATE))
ts_ATM2 <- create_daily_ts(atm_wide$ATM2, min(atm_wide$DATE))
ts_ATM3 <- create_daily_ts(atm_wide$ATM3, min(atm_wide$DATE))
ts_ATM4 <- create_daily_ts(atm_wide$ATM4, min(atm_wide$DATE))
```

By creating a time series object for each ATM machine, we are able to help forecast the information for May 2010

```{r}
tail(atm_wide)
```


Taking a look at the data, we can see that the historical data ends on April 30, 2010 so we need to forecast for the 31 days in May from May 1, 2010 - May 31, 2010. To do this I will be using the ARIMA model for each series, and using the `auto.arima()` function to automatically select an appropriate ARIMA model

```{r}
horizon <- 31

fit_and_forecast <- function(ts_data, h = horizon) {
  model <- auto.arima(ts_data)
  forecast(model, h = h)
}

fc_ATM1 <- fit_and_forecast(ts_ATM1)
fc_ATM2 <- fit_and_forecast(ts_ATM2)
fc_ATM3 <- fit_and_forecast(ts_ATM3)
fc_ATM4 <- fit_and_forecast(ts_ATM4)

par(mfrow = c(2, 2))
plot(fc_ATM1, main = "ATM1 Forecast for May 2010")
plot(fc_ATM2, main = "ATM2 Forecast for May 2010")
plot(fc_ATM3, main = "ATM3 Forecast for May 2010")
plot(fc_ATM4, main = "ATM4 Forecast for May 2010")
par(mfrow = c(1, 1))
```

We fitted an ARIMA model to each ATM's time series and then forecast the next 31 days for the month of May. 

Now we can compile the forecasted dailt value into a data frame and then export then to an excel file as the project outlines

```{r}
end_history <- as.Date("2010-04-30")
forecast_dates <- seq.Date(end_history + 1, by = "day", length.out = horizon)

atm_forecasts <- data.frame(
  Date = forecast_dates,
  ATM1 = as.numeric(fc_ATM1$mean),
  ATM2 = as.numeric(fc_ATM2$mean),
  ATM3 = as.numeric(fc_ATM3$mean),
  ATM4 = as.numeric(fc_ATM4$mean)
)

head(atm_forecasts, 10)
```

Now we can export the forecasts to an excel file

```{r}
write.xlsx(atm_forecasts, file = "ATM_Forecast_May2010.xlsx", sheetName = "Forecast", rowNames = FALSE)
```

```{r}
autoplot(ts_ATM1) +
  autolayer(fc_ATM1, series="Forecast", PI=FALSE) +
  labs(title="ATM1 Historical Data with Forecast", y="Cash (hundreds of dollars)", x="Time") +
  theme_minimal()
```

## Part B - Residential Power FOrecast

```{r}
power_data <- read_excel("ResidentialCustomerForecastLoad-624.xlsx")

head(power_data)
```

Taking a look at this data, we would need to convert the `YYYY-MMM` column into a proper year-month 

```{r}
power_data <- power_data %>%
  mutate(`YYYY-MMM` = as.yearmon(`YYYY-MMM`, format = "%Y-%b"))

str(power_data$`YYYY-MMM`)
head(power_data$`YYYY-MMM`)

head(power_data)
```

Now that we converted our time column into a friendlier format, we can move onto creating a monthly time series

Since the data spans from January 1998 to December 2013, we want to specify a frequency of 12 to indicate that we are dealing with monthly data

```{r}
ts_power <- ts(power_data$KWH, start = c(1998, 1), frequency = 12)

plot(ts_power, main = "Monthly Residential Power Usage from years 1998 - 2013",
     ylab = "KWH", xlab = "Year")
```

For this portion, for model fitting and forecasting the power consumption for 2014, we will also use the `auto.arima()` function to select the best ARIMA model for the historical data and then help us forecast the next 12 months for 2014

```{r}
fit_power <- auto.arima(ts_power)
summary(fit_power)

fc_power <- forecast(fit_power, h = 12)

plot(fc_power, main = "Residential Power Usage Forecast for 2014",
     ylab = "KWH", xlab = "Year")
```

We can see a general seasonality with the forecast of power comsumption for 2014 from the historical data and see some reflections of the sudden dip in poer consumption around 2010-2011

We can now compile the forecasted monthly values into a data frame. TO represent the forecasted months, we can generate a sequence of year mon objects for 2014 and then format them as strings to showcase

```{r}
forecast_months <- seq(as.yearmon("2014-01"), by = 1/12, length.out = 12)

power_forecast_df <- data.frame(
  Month = format(forecast_months, "%Y-%b"),
  Forecast = as.numeric(fc_power$mean)
)

head(power_forecast_df)
```

We can also export the forecasts to an excel file

```{r}
write.xlsx(power_forecast_df, file = "ResidentialPowerForecast_2014.xlsx", sheetName = "Forecast", rowNames = FALSE)
```

## Conclusion

- **Part A - ATM Cash Forecasting**:

For this portion, we were tasked with importing and preprocessing daily ATM transaction data and converting/manipulating the excel file to forecast the cash transaction data for May 2010. We applied to `auto.arima()` function to generate forecasts. Although some ATM showed little seasonality and variability in the data, this may be due to some aspects of the historical data.

- **Part B - Residential Power Usage Forecasting**:

For this portion, we were given an excel sheet containing residential power usage data and we worked with monthly observations which spanned across Janurary 1998 - December 2013. We had to preprocess the data and manipulate the data to get a very time frame column. We created a monthly time series also using the `auto.arima()` method to fit the forecast. We also forecasted for 2014 power usage.

Overall, this project allowed for a comprehensive workflow for time series forecasting. Both aspects of this project portrayed the importance of understanding the data, what we are trying to achieve, and proper preprocessing steps. These forecasts can be useful for further analysis and aid in using other models as well.



